---
title: "ODE Tutorial"
author: "Amy Benefield"
date: "1/28/2020"
output: 
    github_document:
        pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries:
``` {r message=FALSE}
library("deSolve")
```
# 1. A simple ODE: chaos in the atmosphere
We're using the Lorenz equations as the first example. Following along from tutorial (https://cran.r-project.org/web/packages/deSolve/vignettes/deSolve.pdf). The Lorenz equations are 3 DEs (X, Y, Z) representing earth's atmosphere. It's an initial value problem (IVP). First, we must specify the model, and then we implement it.

## 1.1 Model Specification:
Model Specification: define parameters, state variables and their initial values, and writing model equation.
### Setting model parameters, a, b, c:
```{r}
parameters<-c(a=-8/3,
              b=-10,
              c=28)
```
### Setting state variables, X, Y, Z:
```{r}
state<-c(X=1,
         Y=1,
         Z=1)
```
### Defining Model Equations:
Equations are specified in a function that we'll name, Lorenz. Input to function (in order) = model time (t),values of state variables (state), and parameters. Parameter and state variables converted from vectors to a list: with(as.list(c(state, parameters))
```{r}
Lorenz<-function( t, state, parameters){
  with(as.list(c(state, parameters)),{
    # rate of change:
    dX <- a*X + Y*Z
    dY <- b * (Y-Z)
    dZ <- -X*Y + c*Y - Z
    
    # return the rate of change:
    list(c(dX, dY, dZ))
  }) # closing/ending the with(as.list(c(state, parameters)) code
}
```
## 1.2 Model Application:
## Time specification:
Run the model for 100 days, and give output at 0.01 daily intervals.
```{r}
times<-seq(0, 100, by = 0.01)
```
### Model Integration:
Solve the model using the **ode** function from **deSolve** (uses default integration). Function ode takes as input, a.o. the state variable vector (y), the times at which output is required (times), the model function that returns the rate of change (func) and the parameter vector (parms). Function ode returns an object of class deSolve with a matrix that contains the values of the state variables (columns) at the requested output times.
```{r}
out<-ode(y = state, times = times, func = Lorenz, parms = parameters)
head(out)
```
### Plotting Results: 
Use plot method for class of deSolve objects. Outer upper margin increased to allow for figure heading (mtext).
```{r}
par(oma = c(0, 0, 3, 0))
plot(out, xlab = "time", ylab = "-") # Plotted vs. time
plot(out[, "X"], out[, "Z"], pch = ".") # Z vs. X
mtext(outer = TRUE, side = 3, "Lorenz model", cex = 1.5)
```

# 2. Solvers for initial value problems of ordinary differential equations
**deSolve** has several IVP ODE solvers. All integration methods can be used by settting ode's argument method, or can be run as stand alone functions. for each integration routine, several options are available to optimise performance. The next statements will use integration method **radau** to solve the model, and set the tolerances to a higher value than the default. Both statements are the same:
```{r}
outb <- radau(state, times, Lorenz, parameters, atol = 1e-4, rtol = 1e-4)
outc <- ode(state, times, Lorenz, parameters, method = "radau",
            atol = 1e-4, rtol = 1e-4)
```
The default integration method switches automatically between stiff and non-stiff systems. This is a very robust method, but not necessarily the most efficient solver. For most cases, the default solver, **ode** and using the default settings will do. To show how to trigger the various methods, we solve the model with several integration routines, each time printing the time it took (in seconds) to find the solution:
```{r}
print(system.time(out1 <- rk4 (state, times, Lorenz, parameters)))
print(system.time(out2 <- lsode (state, times, Lorenz, parameters)))
print(system.time(out <- lsoda (state, times, Lorenz, parameters)))
print(system.time(out <- lsodes(state, times, Lorenz, parameters)))
print(system.time(out <- daspk (state, times, Lorenz, parameters)))
print(system.time(out <- vode (state, times, Lorenz, parameters)))
```
## 2.1 Runge-Kutta methods and Euler:
RK methods are de novo implementations. They comprise simple Runge-Kutta formulae (Euler’s method euler, Heun’s method rk2, the classical 4th order Runge-Kutta, rk4) and several Runge-Kutta pairs of order 3(2) to order 8(7). ode23 and ode45 are aliases for the popular methods rk23bs resp. rk45dp7. With the following statement all implemented methods are shown:
```{r}
rkMethod()
```
The properties of a Runge-Kutta method can be displayed as follows:
```{r}
rkMethod("rk23")
```
Here **varstep** informs whether the method uses a variable time-step; **FSAL** whether the first same as last strategy is used, while **stage** and **Qerr** give the number of function evaluations needed for one step, and the order of the local truncation error. **A**, **b1**, **b2**, **c** are the coefficients of the Butcher table. Two formulae (rk45dp7, rk45ck) support dense output. It is also possible to modify the parameters of a method (be very careful with this) or define and use a new Runge-Kutta method:
```{r}
func <- function(t, x, parms) {
  with(as.list(c(parms, x)),{
    dP <- a * P - b * C * P
    dC <- b * P * C - c * C
    res <- c(dP, dC)
    list(res)
    })
}

rKnew <- rkMethod(ID = "midpoint",
                  varstep = FALSE,
                  A = c(0, 1/2),
                  b1 = c(0, 1),
                  c = c(0, 1/2),
                  stage = 2,
                  Qerr = 1
                  )

out <- ode(y = c(P = 2, C = 1), times = 0:100, func,
           parms = c(a = 0.1, b = 0.1, c = 0.1), method = rKnew)
head(out)
```

### Fixed time-step methods:
There are two explicit methods that do not adapt the time step: the euler method and the rk4 method. They are implemented in two ways:

* as a **rkMethod** of the general rk solver. In this case the time step used can be specified independently from the times argument, by setting argument **hini**. Function **ode** uses this general code.
* as special solver codes **euler** and **rk4**. These implementations are simplified and with less options to avoid overhead. The timestep used is determined by the time increment in the times argument.

For example, the next two statements both trigger the Euler method, the first using the “special” code with a time step = 1, as imposed by the **times** argument, the second using the generalized method with a time step set by **hini**. Unsurprisingly, the first solution method completely fails (the time step = 1 is much too large for this problem).
```{r}
out <- euler(y = state, times = 0:40, func = Lorenz, parms = parameters)
outb <- ode(y = state, times = 0:40, func = Lorenz, parms = parameters,
            method = "euler", hini = 0.01)
```

## 2.2. Model diagnostics and summaries
Function diagnostics prints several diagnostics of the simulation to the screen. For the Runge-Kutta and lsode routine called above they are:
```{r}
diagnostics(out1)
diagnostics(out2)
```
There is also a summary method for deSolve objects. This is especially handy for multidimensional problems (see below):
```{r}
summary(out1)
```

# 3. Partial differential equations
As package deSolve includes integrators that deal efficiently with arbitrarily sparse and banded Jacobians, it is especially well suited to solve initial value problems resulting from 1, 2 or 3-dimensional partial differential equations (PDE), using the method-of-lines approach. The PDEs are first written as ODEs, using finite differences. This can be efficiently done with functions from R-package **ReacTran**. However, here we will create the finite differences in R-code.
Several special-purpose solvers are included in **deSolve**:

* ode.band integrates 1-dimensional problems comprizing one species,
* ode.1D integrates 1-dimensional problems comprizing one or many species,
* ode.2D integrates 2-dimensional problems,
* ode.3D integrates 3-dimensional problems.

As an example, consider the Aphid model described in Soetaert and Herman (2009). It is a model where aphids slowly diffuse and grow on a row of plants. The model equations are:







